---
output: pdf_document
---
# Vectors
- A vector is an ordered collection of objects of the same type
- The function c(...) concatenates its arguments to form a vector
```{r}
v1 <- c(2.5, 4, 7.3, 0.1)
v1
v2 <- c("A", "B", "C", "D")
v2
v3 <- -3:3
v3
```

# Vectors

- To create a patterned vector:
	- seq() sequence of integers
	- rep() vector of replicated elements
```{r}
seq(0, 2, by=0.5)
seq(0, 2, len=6)
rep(1:5, each=2)
rep(1:5, times=2)
```

# Reference Elements of a Vector
- Use [ ] with a vector/scalar of positions to reference elements of a vector
```{r}
x <- c(4, 7, 2, 10, 1, 0)
x[4]
x[1:3]
x[c(2,5,6)]
x[3] <- 99
x
```

# Reference Elements of a Vector
- Include a minus sign before the vector/scalar to remove elements
```{r}
x[-3]
x[-c(4,5)]
```

# Reference Elements of a Vector
- You can also place conditions inside the brackets
```{r}
x[x > 4]
```

# which() and match()
- Additional functions that will return the indices of a vector
- which() Indices of a logical vector where the condition is TRUE
```{r}
x <- c(4, 7, 2, 10, 1, 0)
x >= 4
which(x >= 4)
```

# which() and match()
- which.max() Location of the (first) maximum element of a numeric vector
```{r}
which.max(x)
x[which.max(x)]
max(x)
```
- which.min() Location of the (first) minimum element of a numeric vector

# which() and match()
- match() First position of an element in a vector
```{r}

y <- rep(1:5, times=5:1)
y
match(1:5, y)
match(unique(y), y)
```

# Vector Operations
- When vectors are used in math expressions the operations are performed element by element
```{r}
x <- c(4,7,2,10,1,0)
y <- x^2 + 1
y
x*y
```

# Useful Vector Functions
Control Structure  | Description
------------- | -------------:
sum(x)/prod(x) | Sum/product of the elements of x
cumsum(x)/cumprod(x) |  Cumulative sum/product of the elements of x
min(x)/max(x)  | Minimum/Maximum element of x
mean(x)/median(x) | Mean/median of x
var(x)/sd(x) | Variance/standard deviation of x
cov(x,y)/cor(x,y) |  Covariance/correlation of x and y 
range(x)  | Range of x
quantile(x) | Quantiles of x for the given probabilities
fivenum(x)  | Five number summary of x
length(x) | Number of elements in x
unique(x) | Unique elements of x
rev(x)  | Reverse the elements of x

# Useful Vector Functions
Control Structure  | Description
------------- | -------------:
sort(x) | Sort the elements of x
which() | Indices of TRUEs in a logical vector
which.max(x)  | Index of the max/min element of x
match() | First position of an element in a vector
union(x, y) | Union of x and y
intersect(x, y) | Intersection of x and y
setdiff(x, y) | Elements of x that are not in y
setequal(x, y)  | Do x and y contain the same elements?

# Vector dimensions

- In R, vectors are *dimensionless*
```{r}
x <- 1:10
dim(x) # Will return NULL
```
- Do not confuse them with column and row vectors in linear algebra; they are neither
  - (But they *can* be cast into one-row or one-column matrices)
  
# Matrices by assigning dimensions

- We can convert a vector to a matrix by assigning dimensions
```{r}
x <- 1:10
dim(x) <- c(5, 2) # 5 rows, 2 columns
x
class(x)
```
- Note that the matrix was filled one column at a time
- (In R, matrices are stored in column-major order)

# Matrices, more directly

- We can also create a matrix more directly (a better way, in most cases)
```{r}
A <- matrix(1:10, nrow=5, ncol=2)
A
```
- Alternatively, we could fill the matrix by row
```{r}
B <- matrix(1:10, nrow=5, ncol=2, byrow=TRUE)
B
```

# Diagonal matrices

```{r}
A <- matrix(c(1, 0, 0, 0, 1, 0, 0, 0, 1), 3, 3)
A
```
A better way:
```{r}
B <- diag(1, 3, 3)
B
```

# Matrix multiplication

- The \* operator is *not* for matrix multiplication
- It is for scalar multiplication (element by element)
```{r}
I2 <- diag(1, 2, 2)
A <- matrix(1:4, 2, 2)
I2 * A
```
- Use %*% instead
```{r}
I2 %*% A
```

# Accessing elements

- To get the $(i, j)$ element of a matrix, use A[i, j], e.g.:
```{r}
A
A[2, 1]
```
- [i] will work as if the matrix were simply a vector
```{r}
A[3]
as.vector(A)
```

# Accessing elements

- Sometimes this is what you want!
```{r}
i <- A > 2
i
A[i]
```
- Or, more directly
```{r}
A[A > 2]
```

# Accessing elements

- Sometimes you'll want an entire row
```{r}
A[1, ]
```
- Or an entire column
```{r}
A[ , 2]
```

# Reference Elements by Name
- We can also reference parts of a matrix by using the row or column names
- Sometimes it is better to reference a row/column by its name rather than by the numeric index. For example, if a program adds or permutes the
columns of a matrix then the numeric index of the columns may change. As a result you might reference the wrong column of the new matrix if you use the old index number. However the name of each column will not change.
- Reference matrix elements using the [ ] but now use the column or row name, with quotations, inplace of the index number

# Reference Elements by Name
- You don't have to specify the names when you create a matrix. To get or set the column, row, or both dimension names of A, use `rownames`, `colnames`, and `dimnames`
- Can also name the elements of a vector, c("name.1"=1, "name.2"=2). Use the function names() to get or set the names of vector elements.

# Reference Elements by Name
```{r}
N <- matrix(c(5,8,3,0,4,1), nrow=2, ncol=3, byrow=TRUE)
colnames(N) <- c("c.1", "c.2", "c.3")
N
N[,"c.2"]
colnames(N)
```

# Reference Elements by Name
```{r}
M  <- diag(2)
(MN <- cbind(M, N))
MN[,2]
MN[,"c.2"]
```

# Useful Matrix Functions
Function  | Description
------------- | -------------:
t(A)  | Transpose of A
det(A)  | Determinate of A
solve(A, b) | Solves the equation Ax=b for x
solve(A)  | Matrix inverse of A
MASS::ginv(A) | Generalized inverse of A (MASS package)
eigen(A)  |  Eigenvalues and eigenvectors of A
chol(A) | Choleski factorization of A
diag(n) | Create an identity matrix
diag(A) | Returns the diagonal elements of a matrix A
diag(x) | Create a diagonal matrix from a vector x
lower.tri(A)/upper.tri(A) | Matrix of logicals indicating lower/upper triangular matrix

# Useful Matrix Functions
Function  | Description
------------- | -------------:
apply() | Apply a function to the margins of a matrix
rbind(...)  | Combines arguments by rows
cbind(...)  |  Combines arguments by columns
dim(A)  | Dimensions of A
nrow(A), ncol(A)  | Number of rows/columns of A
colnames(A), rownames(A)  | Get or set the column/row names of A
dimnames(A) | Get or set the dimension names of A

# Operations across dimensions

- R uses the term *matrix* for two-dimensional data structures, but don't think of them as simply for linear algebra
- Often, you will have data that is naturally in two dimensions, and you'll want to apply some sort of operation across the rows or columns

# Operations across dimensions

- For example, consider 3 drinking water systems that have sampled for a contaminant
- Each system takes 4 samples
- You could organize this into three separate vectors
```{r}
samples1 <- c(0, 0, 2, 0)
samples2 <- c(0, 0, 0, 0)
samples3 <- c(1, 1, 0, 1)
```
- But a matrix is far, far better
```{r}
samples <- matrix(c(samples1, samples2, samples3), 3, 4, byrow=TRUE)
samples
```

# Operations across dimensions

- Now say we want the sum of the results for each system
- Instead of summing the rows individually...
```{r}
sum(samples[1, ])
sum(samples[2, ])
sum(samples[3, ])
```

# Operations across dimensions

- ...you can use the apply function to apply the sum function to each row
```{r}
apply(samples, 1, sum)
```
- The 1 is the dimension to sum over---here, the rows, but we could also use 2 for columns
- Say we wanted to take the mean for each point in time (i.e., for each column)
```{r}
apply(samples, 2, mean)
```

# Example - Simulating Survival Data
- The apply() function is great for simulating survival data. Suppose we want to simulate n = 10 observations where, the event times T follow an exponential distribution with mean λ = 0.25 and the censoring times C are distributed uniformly from 0 to 1. Then the observed data is,
```{r}
n = 10
# Generate event and censor times (look at the documentation to see
# how R parameterizes the exponential distribution)
event  <- rexp(n, 4)
censor <- runif(n)
# Select the minimum time and create an indicator variable
time  <- apply(cbind(censor, event), 1, min)
index <- apply(cbind(censor, event), 1, which.min)-1
cbind(event, censor, time, index)          # Verify
data <- cbind(time, index)                 # Simulated dataset
```

# Arrays
An array is a multi-dimensional generalization of a vector
- To create an array,
```{r}
            array(data = NA, dim = length(data), dimnames = NULL)
```
- Values are entered by columns
- Like with vectors and matrices, when arrays are used in math expressions
the operations are performed element by element.
- Also like vectors and matrices, the elements of an array must all be of the same type (numeric, character, logical, etc.)

# Arrays
```{r}
w <- array(1:12, dim=c(2,3,2),
           dimnames=list(c("A","B"), c("X","Y","Z"), c("N","M")))
w
```

# Reference Elements of an Array
Reference array elements using the [ ] just like with vectors and matrices, but now with more dimensions
```{r}
w <- array(1:12, dim=c(2,3,2),
           dimnames=list(c("A","B"), c("X","Y","Z"), c("N","M")))
w[2,3,1]
w[,"Y",]
w[1,,]
w[1:2,,"M"]
```

# Useful Array Functions
Function  | Description
------------- | -------------:
apply() | Apply a function to the margins of an array
aperm() | Transpose an array by permuting its dimensions
dim(x)  | Dimensions of x
dimnames(x) | Get or set the dimension names of x

# apply()
- We can use the apply() function for more then one dimension
- For a 3-dimensional array there are now three margins to apply the function to: 1=rows, 2=columns, and 3=matrices.
```{r}
apply(w, 2, sum)
apply(w, c(1,3), sum)
```

# Lists
- A list is a general form of a vector, where the elements don't need to be of the same type or dimension.
- The function list(...) creates a list of the arguments
- Arguments have the form name=value. Arguments can be specified with and without names.
```{r}
x <- list(num=c(1,2,3), "Nick", identity=diag(2))
x
```

# Reference Elements of a List
- Elements of a list can be referenced using [ ] as well as [[ ]] or $. 
```{r}
x <- list(num=c(1,2,3), "Nick", identity=diag(2))
x[2]
x[[2]]
x[["num"]]
x$identity
```

# Useful List Functions
Function  | Description
------------- | -------------:
lapply()  | Apply a function to each element of a list, returns a list
sapply()  | Same as lapply(), but returns a vector or matrix by default
vapply()  | Similiar to sapply(), but has a pre-specified type of return value
replicate() | Repeated evaluation of an expression, useful for replicating lists
unlist(x) | Produce a vector of all the components that occur in x
length(x) | Number of objects in x
names(x)  | Names of the objects in x

# Example - lapply()
- First generate a list L of seven different vectors. Then calculate the five number summary of each vector in L using lapply(), sapply() and vapply().
```{r}
# List of three different vectors
L <- lapply(3:5, seq_len)
# Calculate the five number summary for each vector in L
lapply(L, fivenum)
```

# Example - sapply
```{r}
sapply(L, fivenum)
```

# Example - vapply
```{r}
vapply(L, fivenum, c(Min.=0, "1st Quart"=0, Median=0, "3rd Qu"=0, Max.=0))
```

# Dataframes
- R refers to datasets as dataframes
- A dataframe is a matrix-like structure, where the columns can be of different types. You can also think of a dataframe as a list. Each column is an element of the list and each element has the same length.
- A dataframe is the fundamental data structure used by R's statistical modeling functions
- Lecture 8 will be completely devoted to the managment and use of dataframes
```{r}
head(iris)
```

# Numeric
- Technically, numeric data in R can be either double or integer, but in practice numeric data is almost always double (type double refers to real numbers). See ?integer and ?double.
- .Machine outputs numeric characteristics of the machine running R, such as the largest integer or the machine's precision

# Formatting Numbers
- format() formats an object for pretty printing. format() is a generic function that is used with other types of objects. See ?format() for additional arguments.
```{r}
# trim - If FALSE right justified with common width
format(c(1,10,100,1000), trim = FALSE)
format(c(1,10,100,1000), trim = TRUE)
# nsmall - Minimum number of digits to the right of the decimal point
format(13.7, nsmall = 3)
# scientific - Use scientific notation
format(2^16, scientific = TRUE)
```

# Logical

- Logical values are represented by the reserved words TRUE and FALSE in all caps or simply T and F.

Logical  | Description
------------- | -------------:
!x  | NOT x
x&y | x AND y elementwise, returns a vector
x&&y  | x AND y, returns a single value
x\|y | x OR y elementwise returns a vector
x\|\|y  | x OR y returns a single value
xor(x,y)  | Exclusive OR of x and y elementwise
x %in% y  | x IN y

# Logical
Logical  | Description
------------- | -------------:
x\<y | x\<y
x>y | x>y
x <= y | x <= y
x >= y  | x => y
x == y  | x=y
x!=y  | x̸=y

# Logical
Logical  | Description
------------- | -------------:
isTRUE(x) | TRUE if x is TRUE
all(...)  | TRUE if all arguments are TRUE
any(...)  | TRUE if at least one argument is TRUE
identical(x,y)  | Safe and reliable way to test two objects for being exactly equal
all.equal(x,y) | Test if two objects are nearly equal

# Example - Logical Operations
```{r}
x <- 1:10
(x%%2==0) | (x > 5)    # What elements of x are even or greater than 5?
x[(x%%2==0) | (x > 5)]
y <- 5:15              # What elements of x are in y?
x %in% y
x[x %in% y]
```

# Example - Logical Operations
```{r}
any(x>5)
all(x>5)
```


# Isn't that equal?
- In general, logical operators may not produce a single value and may return an NA if an element is NA or NaN.
- If you must get a single TRUE or FALSE, such as with if expressions, you should NOT use == or !=. Unless you are absolutely sure that nothing unusual can happen, you should use the identical() function instead.
- identical() only returns a single logical value, TRUE or FALSE, never NA
```{r}
name <- "Nick"
if(name=="Nick") TRUE else FALSE
# But what if name is never set to "Nick"?

```

```{r, eval=FALSE}
name <- NA
if(name=="Nick") TRUE else FALSE
if(identical(name, "Nick")) TRUE else FALSE
```

# Isn't that equal?
```{r}
(x <- sqrt(2))
x^2
x^2==2
```

# Isn't that equal?
- With all.equal() objects are treated as equal if the only difference is probably the result of inexact floating-point calculations. Returns TRUE if the mean relative difference is less then the specified tolerance.
- all.equal() either returns TRUE or a character string that describes the difference. Therefore, do not use all.equal() directly in if expressions, instead use with isTRUE() or identical().
```{r}
all.equal(x^2, 2)
all.equal(x^2, 1)
isTRUE(all.equal(x^2, 1))
```

# Character
- Character strings are defined by quotation marks (single or double)

Function  | Description
------------- | -------------:
cat() | Concatenate objects and print to console (\n for newline)
paste() | Concatenate objects and return a string
print() | Print an object
substr()  | Extract or replace substrings in a character vector
strtrim() | Trim character vectors to specified display widths
strsplit()  | Split elements of a character vector according to a substring
grep()  | Search for matches to a pattern within a character vector. By default returns a vector of the indices that matched
grepl() | Like grep(), but returns a logical vector
agrep() | Similar to grep(), but searches for approximate matches
regexpr() | Similar to grep(), but returns the position of the first instance of a pattern within a string

# Character
Function  | Description
------------- | -------------:
gsub()  | Replace all occurrences of a pattern with a character vector
sub() | Like gsub(), but only replaces the first occurrence
tolower()/toupper() | Convert to all lower/upper case
noquote() | Print a character vector without quotations
nchar() | Number of characters

# Example - Character Functions

```{r}
animals <- c("bird", "horse", "fish")
home    <- c("tree", "barn", "lake")
length(animals)   # Number of strings
nchar(animals)    # Number of characters in each string
cat("Animals:", animals) # Need \n to move cursor to a newline
cat(animals, home, "\n")
```

# Example - Character Functions
```{r}
paste(animals, collapse=" ")
a.h=paste(animals, home, sep=".") # Pairwise joining of animals and home
# Split strings at ".", fixed=TRUE since "." is used for pattern matching
unlist(strsplit(a.h, ".", fixed=TRUE))
substr(animals, 2, 4) # Get characters 2-4 of each animal
strtrim(animals, 3)   # Print the first three characters
toupper(animals)      # Print animals in all upper case
```

# Example - Pattern Matching
- A regular expression is a pattern that describes a set of strings.
```{r}
# colors() is a character vector of all the built-in color names
colors()[grep("red", colors())]
colors()[grep("^red", colors())]
```

# Example - Pattern Matching
```{r}
colors()[grep("red$", colors())]
colors()[grep("red.", colors())]
```

# Example - Pattern Matching
```{r}
colors()[grep("^[r-t]", colors())] # Colors that begin with r, s, or t
```

# Example - Pattern Matching
```{r}
places <- c("home", "zoo", "school", "work", "park")
gsub("o", "O", places)     # Replace all "o" with "O"
sub("o", "O", places)      # Replace the first "o" with "O"
# Capitalize the first letter, uses Perl-like regular expressions
gsub("(\\w)(\\w*)", "\\U\\1\\L\\2", places, perl=TRUE)
```

# Factors

A factor is a categorical variable with a defined number of ordered or unordered levels. Use the function factor to create a factor variable.
```{r}
factor(rep(1:2, 4), labels=c("trt.1", "trt.2"))
factor(rep(1:3, 4), labels=c("low", "med", "high"), ordered=TRUE)
```

# Factors

Function  | Description
------------- | -------------:
levels(x) | Retrieve or set the levels of x
nlevels(x)  | Return the number of levels of x
relevel(x, ref) | Levels of x are reordered so that the level specified by ref is first
reorder() | Reorders levels based on the values of a second variable
gl()  | Generate factors by specifying the pattern of their levels
  cut(x, breaks)  | Divides the range of x into intervals (factors) determined by breaks

# Example - Factor Functions
```{r}
# Generate factor levels for 3 treatments and 2 cases per treatment
f <- gl(3, 2, labels=paste("trt",1:3, sep="."))
levels(f)
nlevels(f)
relevel(f, "trt.2")
```

# Example - Factor Functions
```{r}
x <- runif(10)
cut(x, 3)
cut(x, c(0,.25,.5,.75,1))
# Cut x into three intervals
# Cut x at the given cut points
```

# Dates and Times
- R has objects that are dates only and objects that are dates and times.
- We will just focus on dates. Look at ?DateTimeClasses for information about how to handles dates and times.
- An R date object has the format: Year-Month-Day
```{r}
Sys.Date()
```

# Dates and Times
Function  | Description
------------- | -------------:
Sys.Date()  | Current date
as.Date() | Cast to date
format.Date() | Format a date
seq.Date()  | Generate a sequence of dates
cut.Date()  | Cut dates into intervals
weekdays, months, quarters  | Extract parts of a date object
julian  | Number of days since an origin

.Date suffix is optional for calling format.Date(), seq.Date() and cut.Date(), but is necessary for viewing the appropriate documentation

# Convert Strings to Date Objects
- Converting a string to a date object requires specifying a format string that defines the date format
- Any character in the format string other then the % symbol is interpreted literally.
- Common conversion specifications (see ?strptime for a complete list),
\%a Abbreviated weekday name %A Full weekday name
\%d Day of the month
\%B Full month name
\%b Abbreviated month name
\%m Numeric month (01-12)
\%y Year without century (be very careful) %Y Year with century
```{r}
dates.1 <- c("5jan2008", "19aug2008", "2feb2009", "29sep2009")
as.Date(dates.1, format="%d%b%Y")
dates.2 <- c("5-1-2008", "19-8-2008", "2-2-2009", "29-9-2009")
as.Date(dates.2, format="%d-%m-%Y")
```

# Sequence of Dates
- To create a sequence of dates,
seq.Date(from, to, by, length.out = NULL)
from, to by
length.out
Start and ending date objects
A character string, containing one of day, week, month, or year. Can optionally be preceded by a (positive or negative) integer and a space, or followed by a . Integer, desired length of the sequence
```{r}
seq.Date(as.Date("2011/1/1"), as.Date("2011/1/31"), by="week")
seq.Date(as.Date("2011/1/1"), as.Date("2011/1/31"), by="3 days")
seq.Date(as.Date("2011/1/1"), by="week", length.out=10)
```

# Cutting Dates
- To divide a sequence of dates in to levels,
```{r, eval= FALSE}
cut.Date(x, breaks, start.on.monday = TRUE)
jan <- seq.Date(as.Date("2011/1/1"), as.Date("2011/1/31"), by="days")
cut(jan, breaks="weeks")
```

# Operations with Dates
- Dates can be added, subtracted, and compared
```{r}
jan1 <- as.Date("2011/1/1")
(jan8 <- jan1 + 7)
jan1 - 14
jan8 - jan1
jan8 > jan1
```

# Formatting Dates
```{r}
format.Date(jan8, "%Y")
format.Date(jan8, "%b-%d")
```

# Missing Data
- R denotes data that is not available by NA
- How a function handles missing data depends on the function. For example
mean only ignores NAs if the argument na.rm=TRUE, whereas which always ignores missing data.
```{r}
x <- c(4, 7, 2, 0, 1, NA)
mean(x)
mean(x, na.rm=TRUE)
which(x>4)
```
- See the documentation for how a particular function handles missing data.

# NULL and NaN
- Quantities that are not a number, such as 0/0, are denoted by NaN. In R
- NaN implies NA (NaN refers to unavailable numeric data and NA refers to any
type of unavailable data)
- Undefined or null objects are denoted in R by NULL. This is useful when we
do not want to add row labels to a matrix. For example,
```{r}
     x <- matrix(1:4, ncol=2, dimnames=list(NULL, c("c.1", "c.2")))
```

# Detecting Missing Data
- To test for missing data avoid using identical() and never use ==. Using identical() relies on unreliable internal computations and == will always evaluate to NA or NaN.
- Functions used for detecting missing data: `is.na`, `is.nan`, and `is.null`
```{r}
x <- c(4, 7, 2, 0, 1, NA)
x==NA
is.na(x)
any(is.na(x))
```

# Detecting Missing Data
```{r}
(y <- x/0)
is.nan(y)
is.na(y)
```

# Testing and Coercing Objects
- All objects in R have a type. We can test the type of an object using a is.type() function.
- We can also attempt to coerce objects of one type to another using a as.type() function.

# Automatic conversions
- Logical values are converted to numbers by setting FALSE as 0 and TRUE as 1
- Logical, numeric, factor and date types are converted to characters by converting each element/level individually

# Some general rules for coercion
- Numeric values are coerced to logical by treating all non-zero values as TRUE
- Numeric characters can be coerced to numbers, but non-numeric characters
cannot
- Factor levels can be coerced to numeric and numbers can be coerced to
factors with a level for each unique number
- Vectors, matrices and arrays are coerced to lists by making each element a
vector of length 1
- Vectors, matrices, arrays can also be coerced from one form to another

# Testing and Coercing Functions
- To test: is.numeric, is.array, is.matrix, is.data.frame, etc.
- To convert: as.numeric, as.array, as.matrix, as.data.frame, etc.

# Example - Testing and Coercing Objects
```{r}
x <- 1:10
x>5
sum(x>5)    # Automatic conversion to numeric vector
is.vector(x)
is.numeric(x)
```

# Example - Testing and Coercing Objects
```{r}
as.list(x)
```

# Example - Testing and Coercing Objects
```{r}
as.numeric("123")
```
