% Advanced Programming

# Classes and Methods

```{r}
species.counts <- table(iris$Species)
print(species.counts)
my.formula <- Petal.Length ~ Petal.Width
print(my.formula)
my.reg <- lm(my.formula, data=iris)
print(my.reg)
```

- How does the `print` function know what to print in each of these cases?

# Classes and Methods

- `print` is actually a *generic method*
- A `print` function is actually defined for each of many different R *classes*
- Think of a class as an extension of data type
- You can define your own classes!

```{r}
class(species.counts)
class(my.formula)
class(my.reg)
```

# Generic Methods

```{r}
print
print.lm
```

# Classes

- Some languages heavily emphasize classes, and, more broadly, object-oriented programming
- R is not like this
- In fact, R's class system is...complicated, with a long and twisted history
- You may encounter the terms *S3*, *S4*, and *reference classes* (or "R5"); these are R's three different class systems
- You will mostly encounter S3 classes, which are simple (good and bad)

# S3 Classes

- Say we have some simulation results that we want to save
- We also want to save the seed used to produce the results
```{r}
seed <- 123
set.seed(seed)
x <- rnorm(10)
my.sim <- list(seed=seed, draws=x)
class(my.sim)
print(my.sim)
```

# S3 Classes

- We could change the class!
```{r}
class(my.sim) <- "sim.res"
class(my.sim)
```

# S3 Classes

- We could even define a print method for this class

```{r}
print.sim.res <- function(sim.res) {
  cat("Seed:", sim.res$seed, "\n")
  cat("Draws:\n")
  print(sim.res$draws)
}
print(my.sim)
```

# Attributes

- Alternatively, we could have stored the seed in an *attribute*

```{r}
my.draws <- x
attr(my.draws, "seed") <- seed
my.draws
```

# Examining structure

```{r}
str(my.sim)
str(my.draws)
```

# Warnings and Errors

- Sometimes (or maybe often) things don't go smoothly

```{r}
1:10 + 1:3
matrix(1:10, ncol=2) %*% matrix(1:3, ncol=3)
```

- *Warnings* are displayed to the user, while *errors* also stop execution

# Generating warnings

```{r}
find.most.freq <- function(x) {
  counts <- table(x)
  names(counts)[which.max(counts)]
}
nums <- c(1, 2, 3, 4, 4, 4, 5, 5, 6)
find.most.freq(nums)
nums <- c(nums, NA)
find.most.freq(nums)
```
- Should we alert the user to the presence of NAs?

# Generating warnings

```{r}
find.most.freq <- function(x) {
  if (any(is.na(x))) {
    warning("x contains NA")
  }
  counts <- table(x)
  names(counts)[which.max(counts)]
}
find.most.freq(nums)
```

# Generating errors

```{r}
find.most.freq <- function(x) {
  if (any(is.na(x))) {
    stop("x contains NA")
  }
  counts <- table(x)
  names(counts)[which.max(counts)]
}
find.most.freq(nums)
```

# Handling warnings

- Just pay attention! R is trying to tell you something important
- But if you have good reason to ignore the warning:
```{r}
suppressWarnings(1:10 + 1:3)
```

# Handling errors

- You can continue execution, essentially ignoring the error:
```{r}
try(matrix(1:10, ncol=2) %*% matrix(1:3, ncol=3))
```

- Or provide some sort of action to take in response:
```{r}
tryCatch(matrix(1:10, ncol=2) %*% matrix(1:3, ncol=3),
         error = function(e) print("Shouldn't have done that!"))
```

# Debugging

- You can stop execution and enter the debugger by calling the `browser` function
- You can also debug an entire function by passing the function to `debug`
- It's possible to locate the source of an error by calling the `traceback` function
- Sometimes, it's easiest to tell R that you want to enter the debugger automatically when an error is thrown:
```{r}
options(error=recover)
```