% Just a quick draft of some array slides
% In Markdown now so I can get something down
% Will transfer to PowerPoint or whatever

# Vector dimensions

- In R, vectors are *dimensionless*
```{r}
x <- 1:10
dim(x) # Will return NULL
```
- Do not confuse them with column and row vectors in linear algebra; they are neither
  - (But they *can* be cast into one-row or one-column matrices)
  
# Matrices by assigning dimensions

- We can convert a vector to a matrix by assigning dimensions
```{r}
x <- 1:10
dim(x) <- c(5, 2) # 5 rows, 2 columns
x
class(x)
```
- Note that the matrix was filled one column at a time
- (In R, matrices are stored in column-major order)

# Matrices, more directly

- We can also create a matrix more directly (a better way, in most cases)
```{r}
A <- matrix(1:10, nrow=5, ncol=2)
A
```
- Alternatively, we could fill the matrix by row
```{r}
B <- matrix(1:10, nrow=5, ncol=2, byrow=TRUE)
B
```

# Diagonal matrices

```{r}
A <- matrix(c(1, 0, 0, 0, 1, 0, 0, 0, 1), 3, 3)
A
```
A better way:
```{r}
B <- diag(1, 3, 3)
B
```

# Matrix multiplication

- The \* operator is *not* for matrix multiplication
- It is for scalar multiplication (element by element)
```{r}
I2 <- diag(1, 2, 2)
A <- matrix(1:4, 2, 2)
I2 * A
```
- Use %*% instead
```{r}
I2 %*% A
```
- Unless you really do want \*
```{r}
3 * I2
```

# Accessing elements

- To get the $(i, j)$ element of a matrix, use A[i, j], e.g.:
```{r}
A
A[2, 1]
```
- [i] will work as if the matrix were simply a vector
```{r}
A[3]
as.vector(A)
```

# Accessing elements

- Sometimes this is what you want!
```{r}
i <- A > 2
i
A[i]
```
- Or, more directly
```{r}
A[A > 2]
```

# Accessing elements

- Sometimes you'll want an entire row
```{r}
A[1, ]
```
- Or an entire column
```{r}
A[ , 2]
```

# Operations across dimensions

- R uses the term *matrix* for two-dimensional data structures, but don't think of them as simply for linear algebra
- Often, you will have data that is naturally in two dimensions, and you'll want to apply some sort of operation across the rows or columns

# Operations across dimensions

- For example, consider 3 drinking water systems that have sampled for a contaminant
- Each system takes 4 samples
- You could organize this into three separate vectors
```{r}
samples1 <- c(0, 0, 2, 0)
samples2 <- c(0, 0, 0, 0)
samples3 <- c(1, 1, 0, 1)
```
- But a matrix is far, far better
```{r}
samples <- matrix(c(samples1, samples2, samples3), 3, 4, byrow=TRUE)
samples
```

# Operations across dimensions

- Now say we want the sum of the results for each system
- Instead of summing the rows individually...
```{r}
sum(samples[1, ])
sum(samples[2, ])
sum(samples[3, ])
```

# Operations across dimensions

- ...you can use the apply function to apply the sum function to each row
```{r}
apply(samples, 1, sum)
```
- The 1 is the dimension to sum over---here, the rows, but we could also use 2 for columns
- Say we wanted to take the mean for each point in time (i.e., for each column)
```{r}
apply(samples, 2, mean)
```

# Operations across dimensions

- Folks with programming backgrounds might understand this with a for loop:
```{r}
result <- vector("numeric", 3) # A vector of length 3 for the results
for (row in 1:nrow(samples)) {
  result[row] <- sum(samples[row, ])
}
result
```
- Or even:
```{r}
result <- vector("numeric", 3)
for (row in 1:nrow(samples)) {
  for (col in 1:ncol(samples)) {
    result[row] <- result[row] + samples[row, col]
  }
}
result
```
- But you usually shouldn't do this! (Plus, isn't it ugly?)

# Modes
- Matrices are not limited to numeric and integer modes
- You can have, for example, a character matrix:
```{r}
x <- matrix(month.name, 3, 4)
x
```

# Arrays

- Why limit ourselves to two dimensions?
- In R, a matrix is a special case of an *array*, which can have many dimensions
- Arrays are quite powerful and allow calculations to be performed quickly

# Arrays

```{r}
x <- array(1:8, dim=c(2, 2, 2))
x
```

# Operations across dimensions

```{r}
apply(x, 1:2, mean)
apply(x, 1, mean)
```