% Basic Programming

# Simulation

- R makes simulating random numbers very easy. For example, from a standard normal:
```{r}
rnorm(5)
```
- Or, say $X \sim \mathcal{N}\left(12, 4\right)$:
```{r}
x <- rnorm(1000000, 12, 2)
mean(x)
var(x)
```
- Note that R parameterizes the normal in terms of the standard deviation

# PDF
-PDF of our example:
```{r}
dnorm(10, 12, 2)
dnorm(12, 12, 2)
dnorm(Inf, 12, 2)
```

# CDF and Quantile Function
- CDF:
```{r}
pnorm(10, 12, 2)
pnorm(12, 12, 2)
pnorm(Inf, 12, 2)
```
- Quantile function:
```{r}
qnorm(0.16, 12, 2)
qnorm(0.5, 12, 2)
qnorm(1, 12, 2)
```

# Distributions

- Type `?Distributions` in R for a list of distributions built-in to R
- Don't see what you need? Try checking CRAN!

# Control Structures

# Coding Practices

- Sometimes you'll just use R interactively, or for very quick scripts
- Other times, you may write longer, more complicated scripts
- It's important to follow some basic best practices so that your code is reliable and can be easily revisited later on

# Best Practices

1. Favor simple, readable solutions
2. Don't repeat yourself
3. Don't re-invent the wheel

# Favor simple, readable solutions

# Don't repeat yourself

# Don't reinvent the wheel