% Data Manipulation
% with the iris dataset

# Motivation

Consider the `iris` dataset.
```{r}
head(iris)
```

# Motivation

We can produce a basic summary of the dataset with `summary`:
```{r}
summary(iris)
```
But what if this particular summary doesn't suit us? What if we're interested in, for example, additional quantiles for the numeric variables? Or what if we want to produce a histogram for each variable?

# A repetitive solution
We could perform operations manually on each variable:
```{r, echo=FALSE}
library(ggplot2)
```
```{r, fig.cap=''}
mean.sepal.length <- mean(iris$Sepal.Length)
sepal.length.lower <- quantile(iris$Sepal.Length, 0.05)
sepal.length.upper <- quantile(iris$Sepal.Length, 0.95)
qplot(Sepal.Length, data=iris, binwidth=0.2)
```
Etc.

# A slightly better solution
But what if we later need to tweak the histogram? Or change the quantiles? We'd have to do this multiple times! A better solution would be to define a function that performs all the needed operations on a parameter. For example:

```{r}
calc.mean.and.quantiles <- function(x) {
  c(mean=mean(x), quantile(x, 0.05), quantile(x, 0.95))
}
calc.mean.and.quantiles(iris$Sepal.Length)
calc.mean.and.quantiles(iris$Sepal.Width)
```

This works fine for small datasets, but what if we don't even know the variable names? Or what if there are too many?

# A more general solution

The data is currently in *wide* form, with a column for each variable. We can put it in *long* form, with a column for variable names and another column for values. The best library for this is `reshape2`.

```{r}
library(reshape2)
iris.long <- melt(iris, "Species")
head(iris.long)
```

# Immediate payoff

```{r, fig.cap=''}
qplot(value, data=iris.long, binwidth=0.2) + facet_wrap(~ variable)
```

# Immediate payoff

```{r, fig.cap=''}
qplot(value, data=iris.long, binwidth=0.2) + facet_grid(Species ~ variable)
```
How many lines would this take if another method were used?

# Summary statistics

```{r}
tapply(iris.long$value, iris.long$variable, function(x) {
  c(mean=mean(x), quantile(x, 0.05), quantile(x, 0.95))
})
```

# Summary Statistics with plyr

```{r}
library(plyr)
ddply(iris.long, .(variable), function(x) {
  c(mean=mean(x$value), quantile(x$value, 0.05), quantile(x$value, 0.95))
})
```

# Summary Statistics with plyr

```{r}
ddply(iris.long, .(variable), summarize,
      mean=mean(value), lower=quantile(value, 0.05), upper=quantile(value, 0.95))
```

# Summary Statistics with plyr

```{r}
ddply(iris.long, .(variable, Species), summarize,
      mean=mean(value), lower=quantile(value, 0.05), upper=quantile(value, 0.95))
```