% Visualization

# Data visualization in R

- R contains many packages for data visualization
- This means that you can usually create the visualization you need, but you may have to sort through packages sometimes
- Fortunately, there are a few basic and powerful graphics packages
- We'll look at `graphics` (included in R) and `ggplot2`

# The graphics package: scatterplots

- R contains a number of easy, built-in functions for graphics
- The simplest is the scatterplot
```{r, fig.cap=''}
x <- seq(from=-10, to=10, length.out=1e2)
y <- x^2
plot(x, y)
```

# Drawing a curve

```{r, fig.cap=''}
plot(x, y, type="l")
```

# Adding labels

```{r, fig.cap=''}
plot(x, y, type="l", col="blue", ylab="x^2")
```

# Adding back the points

```{r, fig.cap=''}
plot(x, y, type="l", col="blue", ylab="x^2")
points(x, y, col="red")
```

# Some real data

```{r}
head(iris)
```

# The graphics package: boxplots

```{r, fig.cap=''}
boxplot(Petal.Length ~ Species, iris)
```

# The graphics package: bar graphs

```{r, fig.cap=''}
barplot(1:3, col=c("red3", "darkgreen", "navyblue"))
```

# ggplot2

- The `ggplot2` library provides an object-oriented, data-centered approach to graphics
- It also produces better-looking graphics by default!
- On the other hand, it takes a bit more work to understand

# Scatterplot

```{r, fig.cap=''}
library(ggplot2)
qplot(Petal.Length, Petal.Width, data=iris)
```

# Adding more dimensions using color

```{r, fig.cap=''}
qplot(Petal.Length, Petal.Width, color=Species, data=iris)
```

# Adding more dimensions using faceting

```{r, fig.cap=''}
qplot(Petal.Length, Petal.Width, data=iris) + facet_wrap(~ Species)
```

# Adding another "geom"

```{r, fig.cap=''}
qplot(Petal.Length, Petal.Width, color=Species, data=iris) + geom_rug()
```

# Adding labels

```{r, fig.cap=''}
qplot(Petal.Length, Petal.Width, color=Species, data=iris) +
  geom_rug() +
  xlab("Petal Length") +
  ylab("Petal Width") +
  ggtitle("My Plot")
```

# Computing statistics: linear regression

```{r, fig.cap=''}
qplot(Petal.Length, Petal.Width, color=Species, data=iris) +
  geom_rug() +
  stat_smooth(method=lm) +
  xlab("Petal Length") +
  ylab("Petal Width") +
  ggtitle("My Plot")
```

# Computing statistics: linear regression

```{r, fig.cap=''}
qplot(Petal.Length, Petal.Width, color=Species, data=iris) +
  geom_rug() +
  stat_smooth(aes(group=0), method=lm) +
  xlab("Petal Length") +
  ylab("Petal Width") +
  ggtitle("My Plot")
```

# Computing statistics: 2D kernel density

```{r, fig.cap=''}
qplot(Petal.Length, Petal.Width, color=Species, data=iris) +
  geom_rug() +
  geom_density2d() +
  xlab("Petal Length") +
  ylab("Petal Width") +
  ggtitle("My Plot")
```

# Another syntax

```{r, fig.cap=''}
ggplot(iris, aes(Petal.Length, Petal.Width, color=Species)) +
  geom_point()
```

# Box plots

```{r, fig.cap=''}
qplot(Species, Petal.Length, data=iris, geom="boxplot")
```

# Histograms

```{r, fig.cap=''}
ggplot(iris, aes(Petal.Length)) +
  geom_histogram(binwidth=0.2) +
  facet_wrap(~ Species, ncol=1)
```